---
title: "soep"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
  pdf_document:
    toc: yes
date: "2023-07-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Include libraries
```{r include=FALSE}
library(tidyverse)
library(ggplot2)
library(stringr)
library(ctsem)
library(ctsemOMX)
```

# Load data

```{r}
soep = readRDS("C:/Users/Linus Widmer/Documents/HU/FM9_Statistics/SOEP-CORE.v37eu_R_EN/R_EN/pl.rds")
biobirth = readRDS("C:/Users/Linus Widmer/Documents/HU/FM9_Statistics/SOEP-CORE.v37eu_R_EN/R_EN/biobirth.rds")
```

# General Data preparation

Filter for BFI items
```{r}
subset_bf <- soep[, c("pid", "syear", "plh0212", "plh0213", "plh0214", "plh0215", "plh0216",
                      "plh0217", "plh0218", "plh0219", "plh0220", "plh0221", 
                      "plh0222", "plh0223", "plh0224", "plh0226")]
```

Remove all participants, that have missing data in their item responses
```{r}
subset_bf_complete = subset_bf[apply(subset_bf, 1, function(x) all(x >= 0)), ]
```

```{r}
nrow(subset_bf) - nrow(subset_bf_complete)
nrow(subset_bf_complete)
```
since the soep dataset is huge and has been assessed since the 90s, a lot of rows get eliminated, as the BFI has first been assessed in 2005.
we are left with 120170 observations. this does not mean we have 120170 participants, since our data is still in wide format and participants have been assessed multiple times.


For the further analysis we will only look at the construct conscientiousness which has 3 items:
- plh0212
- plh0218
- plh0222

In order to calculate the scale we also have to first recode the item plh0218
Calculating the scale
```{r}
subset_bf_complete = as.data.frame(sapply(subset_bf_complete, haven::zap_labels))
subset_bf_complete$plh0218_r = 8 - as.numeric(subset_bf_complete$plh0218)
subset_bf_complete$consc = (subset_bf_complete$plh0212 + subset_bf_complete$plh0222 + subset_bf_complete$plh0218_r) /3
```


```{r}
df_consc_long = subset_bf_complete[, c("pid", "syear", "consc")]
df_consc_long <- df_consc_long %>%
  group_by(pid) %>%
  filter(n() >= 5) %>%
  ungroup()
df_consc_long = merge(df_consc_long, biobirth[, c("pid", "gebjahr")], by = "pid")
df_consc_long$time = df_consc_long$syear - 2005
df_consc_long$min_age = 2005 - df_consc_long$gebjahr
df_consc_long$age = df_consc_long$syear - df_consc_long$gebjahr
```

# Descriptive statistics

mean and standard deviation of consc for every survey year

```{r}
df_consc_long %>%
  group_by(syear) %>%
  summarise(mean_consc = mean(consc),
            sd_consc = sd(consc))
```
mean and standard deviation of conscientiousness for every age group

```{r}
df_consc_long %>%
  group_by(age) %>%
  summarise(mean_consc = mean(consc),
            sd_consc = sd(consc))
```

mean and sd of age per survey year
```{r}
df_consc_long %>%
  group_by(syear) %>%
  summarise(mean_age = mean(age),
            sd_age = sd(age))
```
of course here the mean age for each survey year is rising, as i include only people who have completed all 5 waves. Is this a problem?


Visualization: cross sectional trend of conscientiousness over the life span
```{r}
consc_age_aggregated <- df_consc_long %>%
  group_by(age) %>%
  summarize(mean_value = mean(consc, na.rm = TRUE))

ggplot(consc_age_aggregated, aes(x = age, y = mean_value)) +
  geom_line() +
  xlab("Age") +
  ylab("Conscientiousness") +
  ggtitle("Mean Value Conscientiousness by Age")
```

we can really nicely see how conscientiousness is rising until the 30s and then sort of reaches
a plateau. After 75 years the onservations seem a bit unstable

How many observations per age group
```{r}
hist(df_consc_long$age)
```
```{r}
table(df_consc_long$age)
```

I set the cutoff that n has to be at least 100, meaning i will constrain my dataset
between age ranges 24 and 83


# Continuous time SEM

After some 
deliberations on how to specify the model, we will 
compute models compare them
choose final model
interpret model parameters

for stanct we need the data in a long format, with an id per person and we want to save our
time independent predictor, which will be the mean age of all the assessments a person has 
completed, and our dependent variable consciousness


## ct Model data preparation

adding a quadratic factor and z-standardization
```{r}
df_consc_long$min_agesq = df_consc_long$min_age ^2

df_consc_long$min_age_z = (df_consc_long$min_age - mean(df_consc_long$min_age)) / sd(df_consc_long$min_age)
df_consc_long$min_agesq_z = (df_consc_long$min_agesq - mean(df_consc_long$min_agesq)) / sd(df_consc_long$min_agesq)

```

## Estimating Models

### Linear Model

```{r}
model_ct_consc_TIage_lin <- ctModel(type='stanct', Tpoints = 5,
               LAMBDA = matrix(c(1), nrow = 1, ncol = 1),
               MANIFESTVAR=diag(0, 1), 
               manifestNames = c("consc"),  n.manifest=1,
               id = "pid",
               n.latent=1,
               TRAITVAR = "auto", # accounting for unobserved heterogeneity
               n.TIpred=1, TIpredNames=c("min_age_z"))
model_ct_consc_TIage_lin$pars$indvarying[3] = TRUE # allowing for predictor dependent model coefficients
```

```{r include=FALSE}
fit_ct_consc_TIage_lin = ctStanFit(df_consc_long, model_ct_consc_TIage_lin)
```

## Quadratic Model

Specifying the model
```{r include=FALSE}
model_ct_consc_TIage_sq <- ctModel(type='stanct', Tpoints = 5,
               LAMBDA = matrix(c(1), nrow = 1, ncol = 1),
               MANIFESTVAR=diag(0, 1), 
               manifestNames = c("consc"),  n.manifest=1,
               id = "pid",
               n.latent=1,
               T0VAR = diag(1),
               TRAITVAR = "auto",
               n.TIpred=2, TIpredNames=c("min_age_z", "min_age sq_z"))
model_ct_consc_TIage_sq$pars$indvarying[3] = TRUE # allows for 
```


```{r}
fit_ct_consc_TIage_sq = ctStanFit(df_consc_long, model_ct_consc_TIage_sq)
```
## Model comparison

comparing the linear and quadratic term models using a likelihood ratio test
```{r}
ctChisqTest(fit_ct_consc_TIage_lin, fit_ct_consc_TIage_sq)
```

The chi-sq test of the likelihood of our models is hihgly significant. this means that adding a quadratic term for our time independent predictor fits the data significantly better!
We choose this model now for interpretation of its parameters.


## Interpretation of model parameters

```{r}
summary(fit_ct_consc_TIage_sq)$parmatrices
```

```{r}
summary(fit_ct_consc_TIage_sq, verbose = T)
```


```{r}
ctStanTIpredeffects(fit_ct_consc_TIage_sq, whichpars = c("DRIFT[1,1]", "dtDRIFT[1,1]"), plot = TRUE)
```
here we can see the continuous time (DRIFT) drift coefficient and the discrete time drift coefficient (dtDRIFT).
The dtDrift is just a transformation of the DRIFT coefficient but may be easier to interpret in substantive terms: The lower the dtDRIFT, the higher is the stability of a process and the higher is the autoregresson.
Here we can see that with in increasing age, the dtDRIFT gets smaller and we have a more stable process.
This is inline with the findings of Lucas and Donnellan, who also observed a higher stability in conscientiousness as people get older.

```{r}
ctStanTIpredeffects(fit_ct_consc_TIage_sq, whichpars = c("MANIFESTMEANS[1,1]", "T0MEANS[1,1]"), plot = TRUE)
```
in this plot we can see that with an increasing age also the mean of our dependent variable conscientiousness rises. this is congruent with our descriptive statistics, although i would have expected that there is an asymptotic curve because there is little change of the conscientiousness after the age of 40.

